---
title: "Vaccination Efficiency"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(tidyverse)
library(lubridate)
library(broom)
library(ggthemes)
library(ggrepel)
library(forcats)
```

# Model 1: enzyme kinetics model

# Background

If the rate of new vaccinations per day slows down as more and more people in a country are vaccinated then we can compare the "vaccination efficiency" for each country with analogy to steady-state enzyme kinetics.
The analogy is justifiable since the vaccine infrastructure of a county does not get consumed in converting people from their unvaccinated to their vaccinated state, and each new vaccination is independent of the last vaccination.
If the number of new people vaccinated in a country each day is plotted against the number of unvaccinated people each day, then the curves should look like rectangular parabolas.
Extending the analogy to enzyme kinetics further, the $V_{max}$ would represent the maximum number of people that can be vaccinated (which will be 1) and $K_m$ would represent the catalytic efficiency, i.e. the efficiency of converting unvaccinated people into vaccinated people.

## Unknown

The analogy with enzymes might not work because most countries are far away from reaching the asymptote, i.e. $V_{max}$.

```{r}
covid <- read_excel("owid-covid-data.xlsx")

#just a preliminary plot for each country to check if there are missing values
covid %>% filter(date >= '2020-12-13') %>% 
  ggplot(aes(x = date, y = people_vaccinated_per_hundred, col = location, group = location)) +
  geom_line() +
  theme_classic() +
  theme(legend.position = "none") # legend occupies whole plot so removing it
# instead of above use on the plot labels 

#the above plot revealed that there are missing values, hence the data was cleaned
covid <- covid %>% 
  filter(is.finite(people_vaccinated),
         !is.na(continent)) %>% 
  mutate(date = as_date(date)) %>% 
  mutate(location = as_factor(location))

covid %>% glimpse()
```

```{r}
#checking if the rectangular parabola assumption for applying MM kinetics is met
MM_check <- covid %>% 
  group_by(location) %>% 
  mutate(unvaccinated_people = population - people_fully_vaccinated, 
         new_people_vaccinated_per_day = people_fully_vaccinated - lag(people_vaccinated)
         )

ggplot(data = MM_check, 
       aes(x = unvaccinated_people, 
           y = new_people_vaccinated_per_day,
           # col = location, 
           group = location,)
       ) +
  geom_line() +
  facet_wrap(~ location, 
             scales = "free") + # automatically applies the correct scale to each facet
  coord_fixed() + # maintains shape of plots regardless of shape of output device
  theme_void() + # a completely empty themes that only plots data points
  theme(
    axis.title = element_text(), # adds back the text element removed by theme_void()
    axis.title.y = element_text(angle = 90),
    # aspect.ratio = 1 # fixes the aspect ratio of the whole plot, not just facets
    ) +
  labs(
    x = "Unvaccinated People --->",
    y = "New people vaccinated per day --->")

# fit rectangular parabolas to data and arrange by fit
```

## Conclusion

For countries where a sufficient number of data points are available, the curves fluctuate randomly.

# Model 2: total vaccinations adjusted by population

```{r}
clean_data_1 <- covid %>% 
  group_by(location) %>%
  slice_max(date) #  remove all but one row from each group of rows with 
                  # identical values in a particular column

ranked_1 <- clean_data_1 %>% 
  lm(log(total_vaccinations) ~ log(population), data = .) %>% 
  # glance()
  augment_columns(clean_data_1) %>% 
  arrange(desc(.resid)) %>% 
  head(10)
  
ranked_1 %>% 
  ggplot(aes(x = fct_reorder(location, .std.resid), #reordered location by .std.resid
             y = .std.resid)) +
  geom_point() +
  ylim(0,NA) +
  coord_flip() +
  theme_classic()

clean_data_1 %>% 
  ggplot(aes(y = total_vaccinations, x = population, label = location)) + 
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = 'lm') +
  geom_label_repel(data = ranked_1, #for repelling overlapping text labels away from each other
                 force         = 1,
                 box.padding   = 0.5,
                 point.padding = 0.5,
                 direction     = "x",
                 # nudge_y       = 1e8,
                 alpha = 0.5) +
  theme_classic()

```

# Model 3: people vaccinated per hundred adjusted by population

```{r}
clean_data_2 <- covid %>% 
  group_by(location) %>%
  add_count(location) %>% 
  # arrange(n) %>% 
  filter(n >= 2) %>% 
  nest() %>% 
  mutate(model = map(data, 
                     ~lm(people_vaccinated_per_hundred ~ date, data = .) %>% tidy)
         ) %>% 
  unnest(data) %>% slice_max(date) %>% 
  unnest(model) %>% filter(term == 'date')

ranked_2 <- clean_data_2 %>%
  lm(log(estimate) ~ log(population), data = .) %>% 
  augment_columns(clean_data_2) %>% 
  arrange(desc(.resid)) %>% 
  head(10) 

ranked_2 %>% 
  ggplot(aes(x = fct_reorder(location, .std.resid), #reordered location by .std.resid
             y = .std.resid)) +
  geom_point() +
  ylim(0,NA) +
  coord_flip() +
  theme_classic()

clean_data_2 %>% 
  ggplot(aes(x = population, y = estimate, label = location)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = 'lm') + 
  geom_label_repel(data = ranked_2, #for repelling overlapping text labels away from each other
                 # nudge_y       = 1e8,
                 force         = 1,
                 box.padding   = 0.5,
                 point.padding = 0.5,
                 direction     = "x",
                 alpha = 0.5) + 
  theme_classic()
```
